---
title: "CAN7035 ML/AL Assignment LAQ2"
author: "Ka Yi Chan"
date: "2026-04-19"
output: html_document
---

LAQ 2. (27 marks in total)
All the data required for this LAQ are in ‘For_LAQ2_required_data.zip’

File ‘exp_data_ESCC_90samples.txt’ contains gene expression profiles of 90 oesophageal squamous cell carcinoma samples.

1.	Perform the unsupervised clustering analysis using non-negative matrix factorization (NMF). Find the most appropriate number of clusters, and provide all the essential statistics and plots to support your clusters. (4 marks)

Load packages
```{r}
# Install.packages('NMF')
library(NMF)

# list all available algorithms
nmfAlgorithm()

# retrieve a specific algorithm: 'brunet'
nmfAlgorithm('brunet')
```

Load file ‘exp_data_ESCC_90samples.txt’ contains gene expression profiles of 90 oesophageal squamous cell carcinoma samples.

```{r}
escc_data <- read.table('/Users/kayivionachan/Documents/CAN7035 Machine Learning AI/CAN7035 Assignment/For_LAQ2_required_data/exp_data_ESCC_90samples.txt', header=T, row.names=1)
dim(escc_data)
```

Select the top 1500 most variable genes to reduce noise and focus on genes with the highest variability across samples using mean absolute deviation (MAD).

```{r}
# Feature selection using MAD (Median Absolute Deviation)
escc_exp <- escc_data
mads=apply(escc_exp,1,mad)
escc_exp=escc_exp[rev(order(mads))[1:1500],]
dim(escc_exp)
head(escc_exp)
```

Determine optimal K using NMF rank survey – Test K = 2 to 6, running 10 iterations per rank to evaluate clustering quality. 

```{r}
# Set seed to ensure reproducibility
escc_estim.r <- nmf(escc_exp, 2:6, nrun=10, seed=420)

plot(escc_estim.r)
consensusmap(escc_estim.r)
```

Interpretation of NMK rank survey
Cophenetic correlation: It measures how faithfully the consensus matrix preserves pairwise distances, usually the higher is better. A sharp drop suggests over-splitting. As shown, K = 2 is the highest, however, k = 5 is also good.
Dispersion: higher values indicate more stable/reproducible clusters. K = 2 is the most stable, then sharp drop from K = 2 to 3 then recovered at K = 5.
Silhouette width: higher indicates tighter, better-separated clusters. K = 2 has the best separated clusters, but K = 3, 4 and 5 are also good.K = 3 seems to offer a better balance of high sihouette and high cophenentic correlation.
Sparseness: higher sparseness means metagenes are defined by fewer genes indicates more interpretable. K = 4 and 6 have the highest values, but K. = 5 is also considered as good.

Interpration of consensus map
Consensus map measures how often the same samples appear in the same cluster across multiple different iterations of the NMF algorithm. The colour scale ranges from 0 (blue) to 1 (red). Red means two samples were clustered together in 100% of the runs. Blue means these samples were never clustered together. According to the consensus map, K = 2 has two big red blocks with no noise and K = 3 splits into three distinct blocks but one cluster looks less well-defined. K = 4 shows four reasonably distinct blocks with good separation, its silhouette range 0.99 - 0.5 indicates most samples fit well with a few borderline. K = 5 and 6 clearly show fragmented and many misassigned samples. 

In summary, it seems that K = 2 achieves the highest cophenetic correlation but two clusters is too coarse to capture the molecular heterogeneity within ESCCC. I have chosen K = 4 to be the optimal number of clusters because 1), the cophenetic correlation remains high indicating stable cluster assignments; 2) dispersion shows a recovery at  = 4; 3) sparseness peaks at K = 4 meaning samples are most distinctly assigned to a single cluster and 4) silhouette drops sharply and consensus maps become fragmented beyond K = 4. 

2.	Based on your selection of number of clusters, select and report the most significant gene signatures for each cluster (i.e., top 20 genes based on the differential expression of each cluster against all others). (5 marks)

Perform NMF brunet using k=4, and extract features
```{r}
# Run NMF on escc_exp matrix using k=4 and brunet method
escc_res <- nmf(escc_exp, 4, "brunet", seed=420) 
escc_res

# This function extracts the actual mathematical model (W and H) from the NMF result object, which also contains metadata like the algorithm used and the seed
fit(escc_res)

# This function calculates the product of W x H. In NMF, the goal is for this reconstructed matrix to be as close to the original escc_exp
escc_V.hat <- fitted(escc_res) 
dim(escc_V.hat)

# Quality and performance measures about the factorization are computed by method summary
summary(escc_res)

# Get matrix W
w_matrix <- basis(escc_res)
dim(w_matrix)

# Get matrix H
h_matrix <- coef(escc_res)
dim(h_matrix)
```

NMF reduces the complexity of 90 samples into 4 interpretable biological clusters.
escc_V.hat = 1500 and 90, this confirms NMF successfully reconstructed a martix of the original size.
w_matrix = 1500 and 4, this confirms it has reduced 90 dimensions (samples) down to 4 distinct clusters.
h_matrix = 4 and 90, this confirms it catergorised each of the 90 samples based on how much they express those 4 clusters.

The NMF package implements this procedure in methods featureScore and extractFeature:
```{r}
# Calculate the scores using the Kim-Park method
escc_score <- featureScore(escc_res)
summary(escc_score)

# Extract NMF metagene features
escc_features <- extractFeatures(escc_res)
str(escc_features)

# Identify which are the top genes for each cluster.
# Finding signature genes for Cluster 1
rownames(escc_exp)[escc_features[[1]]]
# Finding signature genes for Cluster 2
rownames(escc_exp)[escc_features[[2]]]
# Finding signature genes for Cluster 3
rownames(escc_exp)[escc_features[[3]]]
# Finding signature genes for Cluster 4
rownames(escc_exp)[escc_features[[4]]]
```

Extract all 4 cluster memberships of NMF clusters. 
IN NMF, every sample has a coefficient score for each of the 4 clusters from the H matrix. The predict function finds which of these clusters has the highest score for each sample, then assign that sample to that specific cluster. 

```{r}
# Extract cluster memberships
cluster_memberships <- predict(escc_res)
table(cluster_memberships)

```

The frequency table shows there are 31 samples in cluster 1, 22 samples in cluster 2, 13 samples in cluster 3 and 24 samples in cluster 4.

Use limma to find significant genes based on differential expression

```{r}
library(limma)

# Create an empty list to store the results when accessing all 4 clusters
top_20_genes_list <- list()
all_genes_list <- list() 

# Create a loop to go through all 4 clusters
for (k in 1:4) {

# Define group and compare one cluster to the rest
  # this creates a temporary label for the 90 samples i.e. the current sample is labeled as ClusterK           if not labeled as Other
  group <- factor(ifelse(cluster_memberships == k, "ClusterK", "Other"))

  # Design matrix. Only calculate the mean expression for ClusterK and Other independently
  mod <- model.matrix(~0 + group)
  # Rename the columns of the matrix
  colnames(mod) <- c("ClusterK", "Other")
  # Align correct sample names to the rows of the design matrix correctly
  rownames(mod) <- colnames(escc_exp)
  
  # Fit a linear model for all 1500 genes
  fit <- lmFit(escc_exp, mod)
  # Contrast - Subtract the average of Other from the average of ClusterK
  con.matrix <- makeContrasts(ClusterKvsOther = ClusterK - Other, levels = mod)
  # Apply contrast to the linear model
  fit2 <- contrasts.fit(fit, con.matrix)
  fit2 <- eBayes(fit2)
  
  # Top 20 DE genes
  top20 <- topTable(fit2, adjust = "BH", number = 20, sort.by = "logFC")
  top_20_genes_list[[k]] <- top20
  
  # All DE genes for GSEA
  all_genes <- topTable(fit2, adjust = "BH", number = Inf)
  all_genes_list <- append(all_genes_list, list(all_genes))
  
  cat("\n=== Top 20 DE Genes for Cluster", k, "vs all others ===\n")
  print(top20)
}
```

Based on the 20 most differentiated gene expression signatures across these four clusters, we can infer the corresponding cellular subtypes.

Cluster 1 indicates a highly proliferative squamous-cell subtype, supported by the upregulation of COL17A1, KRT16, and KRT6B, which are markers of hyper-proliferative squamous epithelium. Downregulation of KRT4 indicates a significant dedifferentiation, where cells lose their mature squamous identity to favor a more proliferative state. The loss of ALDH1A1 enzyme responsible for cellular detoxifying and managing oxidative stress a metabolic shift that often seen in poorly differentiated ESCC tumors. 

Cluster 2 likely belongs to a mesenchymal/stroma-rich subtype, as desmin (DES) a structural protein is upregulated, while squamous keratins (KRT16 and KRT6C) and Small Proline-Rich (SPRR) protein family (SPRR1B, SPRR2E) and C10orf99 are downregulated. This combination indicates a total loss of the epithelial barrier function in favor of a more invasive cellular architecture, often associated with metastasis in ESCC.

Cluster 3 identifies well-differentiated subtype, this signature is defined by the upregulation of mucodal keratins (KRT4 and KRT13) and Small Proline-Rich Proteins (SPRR3 and SPRR2A) High expression of CLCA4 further confirms a mature cellular state.

Cluster 4 belongs to a metabolic-like subtype, since genes associated with detoxification and metabolism enzymes such as ALDH3A1, ADH7, AKR1C3, CES1 and members of the Cytochrome P450 family (CYP4F11 and CYP4F3) are all upregulated.

3.	Perform GSEA and report the most upregulated and downregulated pathways for each cluster, and comment on your findings. (5 marks)

```{r}
# load packages
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(pathview)
library(DOSE)

# Prepare ranked gene lists from Q2 limma results
gene_lists <- list()

for (i in 1:4) {
  gene_list <- all_genes_list[[i]]$logFC # logFC from Q2 limma
  gene_list <- log2(10^gene_list) # convert to log2FC
  names(gene_list) <- rownames(all_genes_list[[i]]) # gene names from Q2 limma
  gene_list <- sort(gene_list, decreasing = TRUE)
  gene_lists[[i]] <- gene_list # stores directly into gene_lists
}

# Run GSEA 
GSEA_res <- list()

for (i in 1:4) {
  set.seed(420)
  gse <- gseGO(geneList = gene_lists[[i]], 
               ont = "ALL", 
               keyType = "SYMBOL", 
               nPerm = 1000, 
               minGSSize = 3, 
               maxGSSize = 800, 
               pvalueCutoff = 0.05, 
               verbose = TRUE, 
               OrgDb = "org.Hs.eg.db", 
               pAdjustMethod = "none")
  GSEA_res[[i]] <- gse 
}

# Dotplots 
for (i in 1:4) {
  plot(dotplot(GSEA_res[[i]], showCategory = 10, split = ".sign",
               title = paste0("Cluster ", i), font.size = 6) + 
         facet_grid(. ~ .sign))
}

```

Interpretation of dotplots
The name of the pathways are on the y-axis, while the plots are divided into activated (upregulated) and suppressed (downregulated) categories for each cluster. The count and bubble size represent the number of genes from each specific pathway identified in the dataset. The color of the p.adjust bubbles represents statistical significance, ranging from red (more significant) to blue (less significant).  

Cluster 1
The activated pathways in Cluster 1 involve long-chain fatty acid and transmembrane transporter activity, suggesting that these tumors have increased activity in lipid transport and membrane remodeling. The previously observed upregulation of COL17A1 (a structural collagen) and KRT16 (an epithelial filament protein) suggests that these structural keratins work in tandem with membrane transporters to maintain the hyper-proliferative state of squamous cells.

Conversely, the suppressed pathways include lipid metabolic processes, carboxylic acid metabolic processes, and neuron projection/synapse pathways. This suggests a significant shift in the cell composition and metabolic streamlining within the tumor microenvironment of Cluster 1. 

Cluster 2
Cluster 2 shows activation of platelet-derived growth, extracellular matrix (ECM) structural constituents, and MHC class II protein complex assembly. This confirms a "remodeling" phenotype, as these tumor cells are producing significant connective tissue and undergoing Epithelial-Mesenchymal Transition (EMT). Furthermore, the upregulation of MHC Class II and peptide antigen assembly suggests an immunologically "hot" tumor microenvironment, characterized by high infiltration of antigen-presenting cells into the tumor stroma.

The suppressed pathways, including tissue, epithelial, and epidermis development, indicate that Cluster 2 samples have lost their squamous identity and shifted toward a more invasive, mesenchymal state.

Cluster 3
The activated pathways in Cluster 3 are dominated by keratinization, epidermal cell differentiation, and skin/epithelium development. The presence of SPRR (small proline-rich proteins) and KRT78 among the top 20 DE genes suggests that these cells are in a state of late-stage squamous differentiation.

The suppressed pathways include the regulation of collagen fibril organization, endodermal cell differentiation, and various collagen trimers. Cluster 3 shows significantly less basement membrane remodeling activity compared to Clusters 1 and 2, suggesting this is a well-differentiated molecular subtype that maintains the structural characteristics of normal esophageal mucosa.

Clsuter 4
Cluster 4 is the most metabolically active group, with activated pathways including lipid, organic acid, and monocarboxylic acid metabolic processes, as well as oxidoreductase activity. This aligns with the Q2 findings where ALDH1A1 and UGT1A1—genes responsible for detoxification and metabolic turnover were highly upregulated.

In contrast to the "hot" profile of Cluster 2, Cluster 4 suppresses MHC class II protein complex assembly and Toll-like receptor signaling. This suggests an immunologically "cold" phenotype, as the tumor cells appear to be actively evading immune detection.

In summary, GSEA analysis is able to classify:
Cluster 1 as the proliferative subtype 
Cluster 2 as the mesenchymal/infammed subtype
Cluster 3 as the differentiated subtype
Cluster 4 as the metabolic/immune evasive subtype.

4.	Repeat the procedure 1-3 above using the consensus clustering algorithm (e.g.,ConsensusClusterPlus R package) and report all the required findings. (5 marks)

Data normalisation prior to ConsensusClusterPlus
```{r}
mads=apply(escc_data,1,mad)
escc_data2=escc_data[rev(order(mads))[1:1500],]

# Data is normalized by centering around gene median
c_escc_data = sweep(escc_data2,1, apply(escc_data2,1,median))

```

```{r}
library(ConsensusClusterPlus)

consensus_escc = ConsensusClusterPlus(as.matrix(c_escc_data), 
                                      maxK=10, # Run 10 clusters
                                      reps=500, # 500 repetitions
                                      pItem=0.8,
                                      pFeature=1,
                                      clusterAlg="hc",
                                      distance="pearson",
                                      seed=420,
                                      plot=NULL)
```

According to the Delta Area plot, the "Rule of Thumb" is to find the K value where the relative increase in the area under the CDF curve drops off significantly. The plot clearly shows while the maximum relative change in the area under the CDF curve occurs at K = 2 and K =3, these values are nearly identical, suggesting a plateau of stability. A significant drop occurs at K = 4, after which the curve flattens toward zero. This indicates that K = 4 is the optimal number of clusters, capturing the most complex biological structure before reaching a point of diminishing returns in cluster stability.

Extract the results from ConsensusClusterPlus object
```{r}
escc_clusters <- consensus_escc[[4]]$consensusClass
table(escc_clusters)
```

Use limma to find significant genes based on differential expression

```{r}
# Create an empty list to store the results when accessing all 4 clusters
cc_top_20_list <- list()
cc_all_genes_list <- list()

# Create a loop to go through all 4 clusters
for (k in 1:4) {

  cc_group <- factor(ifelse(escc_clusters == k, "ClusterCC", "Rest"))
  
  # Design matrix
  cc_mod <- model.matrix(~0 + cc_group)
  colnames(cc_mod) <- c("ClusterCC", "Rest")
  rownames(cc_mod) <- colnames(escc_exp)
  
  # Fit linear model
  cc_fit <- lmFit(escc_exp, cc_mod)
  # Contrast - Subtract the average of Other from the average of ClusterK
  cc_con.matrix <- makeContrasts(ClusterCCvsRest = ClusterCC - Rest, levels = cc_mod)
  # Apply contrast to the linear model
  cc_fit2 <- contrasts.fit(cc_fit, cc_con.matrix)
  cc_fit2 <- eBayes(cc_fit2)
  
  # Top 20 DE genes
  cc_top20 <- topTable(cc_fit2, adjust = "BH", number = 20, sort.by = "logFC")
  cc_top_20_list[[k]] <- cc_top20
  
  # All DE genes for GSEA
  cc_all_genes <- topTable(cc_fit2, adjust ="BH", number =Inf)
  cc_all_genes_list <- append(cc_all_genes_list, list(cc_all_genes))
  
  cat("\n=== Top 20 DE Genes for Consensus Clustering", k, "vs all others ===\n")
  print(cc_top20)
}
```

Cluster 1 indicates a well-Differentiated subtype due to high expression of genes associated with epidermal differentiation complex, such as CRCT1, CNFN, and the SPRR family (SPRR2E, SPRR3, SPRR2A). This signature correspond to the signature identified in NMF cluster 3. 

Cluster 2 indicates a metabolic/detoxification subtype because of high level expression of ALDH3A1 & AKR1C3/AKR1B10 which are the cell's primary defense against oxidative stress and toxic byproducts of metabolism; CES1 which involved in the hydrolysis of various xenobiotics (foreign substances like drugs or toxins and CYP4F11 which is a Cytochrome P450 enzyme involved in the metabolism of lipids and fatty acids.

Cluster 3 represents an undifferentiated or basal-like subtype because there are significant downregulation of squamous maturation markers, including the Small Proline-Rich Protein (SPRR) family and peptidase inhibitor (PI3). PI3 stops enzymes from breaking down the epithelial structure, hence loss of PI3 lead to tumour microenvironment more susceptible in structural breakdown. 

Cluster 4 is identified as the inflamed subtype as shown by upregulation of immune-effector genes (CXCL10, GBP5 & GNLY) indicating a T-cell infiltration and interferon signaling are taking place. This cluster also shows loss of squamous and metabolic markers such as AKR1C2, CES1 and SPRR3 suggesting that the biological profile is shifted entirely toward an immune-active state.

Perform GSEA and report the most upregulated and downregulated pathways for each cluster

```{r}
# Perform ranked gene lists from limma results
cc_gene_lists <- list()

for (i in 1:4) {
  cc_gene_list <- cc_all_genes_list[[i]]$logFC
  cc_gene_list <- log2(10^cc_gene_list)
  names(cc_gene_list) <- rownames(cc_all_genes_list[[i]])
  cc_gene_list <- sort(cc_gene_list, decreasing = TRUE)
  cc_gene_lists[[i]] <- cc_gene_list
}

# Run GSEA 
cc_GSEA_res <- list()

for (i in 1:4) {
  set.seed(420)
  cc_gse <- gseGO(geneList = cc_gene_lists[[i]], 
               ont = "ALL", 
               keyType = "SYMBOL", 
               nPerm = 1000, 
               minGSSize = 3, 
               maxGSSize = 800, 
               pvalueCutoff = 0.05, 
               verbose = TRUE, 
               OrgDb = "org.Hs.eg.db", 
               pAdjustMethod = "none")
  cc_GSEA_res[[i]] <- cc_gse
}

# Dotplots 
for (i in 1:4) {
  plot(dotplot(cc_GSEA_res[[i]], showCategory = 10, split = ".sign",
               title = paste0("Cluster ", i), font.size = 6) + 
         facet_grid(. ~ .sign))
}
```

Interpretation of dotplots (ConsensusClusterPlus)
Cluster 1 
It shows activation of epithelial cell differentiation, secretory granule, secretory vesicle, cytosol, membrane and extracellular space and region. These suggests tumours retained epithelial function. 

The suppressed pathways include 1) structural/extracellular matrix pathways e.g. elastic fiber assembly, elastic fiber, aorta morphogenesis; 2) growth signalling pathway e.g. platelet-derived growth factor binding; 3) tissue remodelling e.g. intramembranous ossification, direct ossification and 4) neural/stromal microenvironment pathways e.g. perisynaptic/synapse-associated extracellular matrix. Suppression of these pathways suggest reduced stromal remodelling. 

Cluster 2 
It is characterised as metabolically active tumours with with immune suppression as it has a strong activation of metabolic pathways e.g. primary metabolic process, small molecule metabolic process, lipid/carboxylic/organic acid metabolic process. It also shows high oxidative activity.

Immune-related pathways are suppressed in cluster 2, particularly MHC class II antigen presentation and innate immune response. This indicates a cold tumour with potential immune evasion.

Cluster 3 
It shows strong activation of stromal and mesenchymal pathways e.g. aorta smooth muscle tissue morphogenesis, fibroblast migration and cell adhesion pathway e.g. substrate adhesion-dependent cell spreading. This cluster also show increased immune-related activity such as immunoglobulin receptor binding, complement-dependent cytotoxicity. 

Conversely, the suppressed pathways include epidermis development, epithelial cell differentiation, epithelium development, metabolic activity and developmental pathways. This profile suggest a mesenchymal/stromal-enriched tumour subtype potentially reflecting epithelial-to-mesenchymal transition (EMT).

Cluster 4 
It shows broad activation of general cell signalling, communication, and stress response pathways e.g. positive regulation of biological process, regulation of response to stimulus, signal transduction, signalling, cell communication. 

Detoxification and drug metabolism pathways are strongly suppressed including glutathione metabolism, redox homeostasis, and daunorubicin/doxorubicin metabolism. This suggests these tumours may have reduced drug detoxification capacity, potentially making them more sensitive to chemotherapy. 

In summary, GSEA analysis identified these 4 clusters generated by ConsensusClusterPlus as:
Cluster 1 belongs to the Well-Differentiated (Epithelial-Stable) Subtype.
Cluster 2 belongs to the Metabolic / Immune-Cold Subtype
Cluster 3 belongs to the Mesenchymal / Immune-Hot Subtype
Cluster 4 belongs to the Stress-Responsive / Chemo-Sensitive Subtype
 
5.	Finally, compare the clusters and cluster membership between that based on NMF and that from consensus clustering. Are these clusters similar? Are the upregulated and downregulated pathways identified from the two algorithms similar? Comment on the comparison. (4 marks)

Subtype profile: Well-differentiated
NMF cluster 3
Consensus cluster 1
Both algorithms show epithelial differentiation and identified KRT4/13 and SPRR genes.

Subtype profile: Metabolic / Immune-Cold
NMF cluster 4
Consensus cluster 2
Both algorithms show metabolic activation and immune suppression. 

Subtype profile: Mesenchymal / Invasive
NMF cluster 2
Consensus cluster 3
Both algorithms show stromal/mesenchymal activation and immune inflamed

Subtype profile: Aggresive/Stress
NMF cluster 1
Consensus cluster 4
Both algorithms show active signalling, stress response with NMF focused on the proliferative side while consensus focused on the stress and chemosensitive side.








